---
output:
  pdf_document: default
  html_document: default
---
# Descriptive statistics

## Quick summary statistics
The first step of descriptive statistics is to have a **quick (smart) overview** of the data. And always ask yourself **"is the output consistent with what you expected in terms of the variables’ types?"**

::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}
```{r, echo=F, eval=T}
load("sleepApnea-clean.Rdata")
```

To have a descriptive summary, the classical R function is *summary()*
```{r, eval=F}
summary(sleep[, 1:3])
```

Note that in the example above we asked for the summary of the first 3 columns of the dataset. The indexing of data set in a matrix-like format often become useful and is similar to playing battle ship. You need to use **dataset[row_index,column_index]** to filter on rows, before the comma,  or on columns, after the comma.

Additionnal statistical summary functions can be find in the other R libraries, like in the package *pastecs* (function stat.desc()).
:::

:::{}
### Stata{-}
The *codebook* function returns the main characteristics of the data set, *i.e.* the number of observations, number of variables, and characteristics of variables.
```{stata, eval=F}
codebook
```
:::
::::

## Central and deviation parameters
::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}
```{r, eval=F}
mean(sleep$age)
sd(sleep$age)
median(sleep$age)
quantile(sleep$age, probs=c(0.25, 0.75))
```
:::

:::{}
### Stata{-}

```{stata, eval=F}
summarize age, detail
````

:::
::::

## Confidence interval
::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}
```{r, eval=F}
library(epiDisplay)
ci(sleep$age)
```
:::

:::{}
### Stata{-}
:::
::::

## Cross table
::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}
```{r}
crosstab <- table(sleep$gender,sleep$diabetes)
round(prop.table(crosstab,1)*100,digits=2)
round(prop.table(crosstab,2)*100,digits=2)
```
:::

:::{}
### Stata{-}
```{stata, eval=FALSE}
tabulate gender
```
:::
::::