---
output:
  pdf_document: default
  html_document: default
---
# Descriptive statistics

## Quick summary statistics
The first step of descriptive statistics is to have a **quick (smart) overview** of the data. And always ask yourself **"is the output consistent with what you expected in terms of the variables’ types?"**

::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}
```{r, echo=F, eval=T}
load("sleepApnea-clean.Rdata")
```

To have a descriptive summary, the classical R function is *summary()*
```{r, eval=F}
summary(sleep[, 1:3])
```

Note that in the example above we asked for the summary of the first 3 columns of the dataset. The indexing of data set in a matrix-like format often become useful and is similar to playing battle ship. You need to use **dataset[row_index,column_index]** to filter on rows, before the comma,  or on columns, after the comma.

Additionnal statistical summary functions can be find in the other R libraries, like in the package *pastecs* (function stat.desc()).
:::

:::{}
### Stata{-}
The *codebook* function returns the main characteristics of the data set, *i.e.* the number of observations, number of variables, and characteristics of variables.
```{stata, eval=F}
codebook
```
:::
::::

## Central and deviation parameters

Central and deviation parameters are essential parameters to summaryze and describe the distribution of your data. The most well not central parameter is the mean but you sholud know that it is not the only one and not always the best one to use, especially when you have skewed distribution. These parameters should be look at together and with the help of graphical representation for better description.

::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}

In **R**, you sholud think english what is the term for average it is "mean" and it is the name of the function;  same for median and quantile. If you do not know the mane of the function search in the help pages.
```{r, eval=F}
# average
mean(sleep$age)
# variance
var(sleep$age)
# standard deviation
sd(sleep$age)
# median 
# When sort in increasing order, 50% of values below and above
median(sleep$age)
# interquartile range
IQR(sleep$age)
```

For the *quantile()* the *probs=* argument allows you to specify which quantile of the distribution you want to look at. 
```{r, eval=T}
# 25%-75%
quantile(sleep$age, probs=c(0.25, 0.75))
#quartile
quantile(sleep$age, probs=seq(0, 1, 1/4))
# tercile
quantile(sleep$age, probs = seq(0,1,1/3))
```
**Interpretation:** For the age, the 1$^{st}$ quantile indicates that 25% of the patients in ou sample are 42 years old or younger, wile 75% (3$^{rd}$ quantile ) are 54 years old or older.
:::

:::{}
### Stata{-}

```{stata, eval=F}
summarize age, detail
````

:::
::::

## Confidence interval {#confidence-interval}

The confidence interval represents the interval in which the true value of the estimated parameter stands with a certain confidence (often 95% - allowing 5% of the estimation being wrong if 100 estimation were perfomed).

::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}
In **R** the *epiDisplay* libary is useful for many computations in epidemiology and notably the confidence interval of the mean. A simple call to the ci() function will compute the 95% confidence interval. The level of confidence is specify using the risk alpha conterpart , here the default is alpha=0.05 (so not need to specify it). 
```{r, eval=T, message=FALSE, warning=FALSE}
library(epiDisplay)
ci(sleep$age)
ci(sleep$age, alpha = 0.01)
```
:::

:::{}
### Stata{-}
In **Stata**, the syntax is close but the parameter to estimate must be specified and the level of confidence is specified in term of confidence.
```{stata, eval=F}
ci means age
ci means age, level(99)
```
:::
::::

## Cross table
Contingency table or cross tabulated table are needed to compute absolute or relative frequency (proportion) of categorical variables.

::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}
In **R** the basic function are *table()* and *prop.table()*. 
```{r}
crosstab <- table(sleep$gender,sleep$diabetes)
```

In *prop.table()* the  margin argument specify the dimension (row or column) along which the relative frequencies are computed.
```{r}
# The 100% are on the grand total
round(prop.table(crosstab)*100,digits=2)
# Percentage by row
round(prop.table(crosstab,margin=1)*100,digits=2)
# Percentage by column
round(prop.table(crosstab,2)*100,digits=2)
```
:::

:::{}
### Stata{-}
In **Stata** the *tabulate* function produces a two-way table of frequency counts.
```{stata, eval=FALSE}
tabulate gender diabetes
tabulate gender diabetes, row
tabulate gender diabetes, column
```
The *tabulate* function can be used along with various measures of association, including the common Pearson’s Chi-square and Fisher’s exact test (See [univariate test](Univariate statistical test) for examples).
:::
::::