---
output:
  html_document: default
  pdf_document: default
---

# Plotting

Visualization are often of great help to explore and summarize your data. **The plots should be chosen wisely and according to your data type** (*i.e.* continuous, categorical). *Stata* presents rather basic graphical functions but *R* possess extensive libraries for a wide range of visual displays going from the basic we will cover here to some geomatic representations (ex library *mapsf*). A quick introduction to the *ggplot2* library (or grammar for graphic) is done here too but for more ressources see [https://ggplot2.tidyverse.org/](https://ggplot2.tidyverse.org/) 

## Quantitative data

For **one quantitative continuous variable**, the **histogram** is the plot to use to visualize the global trend of the data. However you need to be aware that most statistical software optimize the bin width to display the values per intervals but you may change it and you may seen that it change the visual and possibly your interpretation.

::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}
```{r, echo=FALSE}
load("sleepApnea-clean.Rdata")
```

In **R**, you can specific the number of breaks
```{r, eval=FALSE}
hist(sleep$age)
```

```{r histogram, eval=TRUE, fig.cap='Distribution of age (in years)', out.width='70%', fig.asp=.75, echo=FALSE, message=FALSE, warning=FALSE}
hist(sleep$age, breaks=5, 
     main="Distribution of age",
     xlab="age")
```

A more advanced graph with *ggplot2* library, where *aes()* is the aesthetics or the dimensions along with you plot the data and the *geom\*\_()*, the geometry you give to the plot. You may have noticed that the using ggplot is more like writing a sentence with `+` sign to delimit the instructions.
```{r, eval=TRUE}
library(ggplot2)
ggplot(sleep, aes(age)) + 
  geom_histogram(binwidth = 5,  boundary=TRUE,
                 col="grey", fill="lightblue") +
  ggtitle("Distribution of age")
```
:::

:::{}
### Stata{-}

In **Stata**, the width argument allows you to specify the interval range
```{stata, eval=FALSE}
//histogram of age with two year intervals
hist age, width(2)
//histogram of age with five year intervals
hist age, width(5)
```

:::
::::

For **one quantitative continuous variable group by a categorical value**, the **Boxplot** is especially useful for comparison.

::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}
With **R** the syntax is:
```{r, eval=TRUE}
boxplot(age ~ gender , data=sleep, 
     main="Distribution of age \ by gender",
     xlab="gender",
     col="lightblue")
```

Note:  the `~` sign that is the symbol for the formula type syntax of Y function of X. You will see again that notation when performing statistical test and regression modeling. 

With *ggplot2* you could do
```{r, eval=TRUE}
ggplot(sleep, aes(y=age, x=gender)) + 
  geom_boxplot(fill="lightblue") +
  ggtitle("Distribution of age by gender")
```
:::

:::{}
### Stata{-}
In **Stata**, the grouping variable is called by the over() option.
```{stata, eval=FALSE}
graph box age, over(gender)
```
:::
::::

When looking for (linear) relationship between **two quantitative continuous variable**, you want to do a **scatterplot**.

::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}
With **R** the syntax is:
```{r, eval=TRUE}
plot(sleep$bmi , sleep$SleepApnea, 
     main="BMI function of age",
     xlab="age", ylab="bmi")
```

With *ggplot2*
```{r, eval=TRUE}
ggplot(sleep, aes(bmi , SleepApnea)) + geom_point()+
  ggtitle("BMI function of age")
```
:::

:::{}
### Stata{-}
In **Stata**, the grouping variable is called by the over() option.
```{stata, eval=FALSE}
scatter SleepApnea bmi
```
:::
::::

## Qualitative data (categorical)

For **one categorical variable** by group or not you should use a **barplot**.

::::{style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 30px;"}
:::{}
### R{-}
```{r, eval=TRUE}
db2gender <- table(sleep$diabetes, sleep$gender)
db2gender_prop <- round(prop.table(db2gender,margin=2)*100,2)
barplot(db2gender, 
     main="Repartition of diabetes",
     xlab="gender")
barplot(db2gender_prop, 
     main="Proportion of diabetes by gender",
     xlab="gender",
     ylab="Proportion", 
     legend.text = c("No", "Yes"))
```

With *ggplot2* you could do
```{r, eval=TRUE}
ggplot(sleep, aes(diabetes, fill=gender)) + 
  geom_bar() +
  ggtitle("Distribution of age by gender")
```

:::

:::{}
### Stata{-}
```{stata, eval=FALSE}
graph bar diabetes, over(gender)
```
:::
::::

